---
title: "Investigate correlation in z scores "
author: "Matthew Stephens"
date: 2017-03-03
output: html_document
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

```{r knitr-opts-chunk, include=FALSE}
```

**Last updated:** `r Sys.Date()`

**Code version:** `r workflowr::extract_commit(".", 1)$sha1`


# Correlation of "nullish" z scores

Read in $z$ scores and look at correlations between "nullish" (very non-signif) $z$ scores.
In this case I define nullish as absolute value <2.
```{r}
source("../code/SumstatQuery.R")
  ##
  # Load data for a random subset of eqtls
  dat <- GetSS("null", "../data/MatrixEQTLSumStats.Portable.h5")
  bc = "Brain_Cerebellum"
  bch = "Brain_Cerebellar_Hemisphere"
 
  plot(dat$z[,bc],dat$z[,bch])
  nullish = abs(dat$z[,bc])<2 & abs(dat$z[,bch])<2
  cor(dat$z[nullish,bc],dat$z[nullish,bch])
  
```


Compare whole blood and cerebellum:
```{r}
  wb = "Whole_Blood"
  plot(dat$z[,bc],dat$z[,wb])
  nullish = abs(dat$z[,wb])<2 & abs(dat$z[,bc])<2
  cor(dat$z[nullish,wb],dat$z[nullish,bc])
  
```


# Difference in Z scores

Assume $$z^r_j = \mu^r_j + e_j$$ where $e_1,e_2$ are joint normal with variance 1 and some covariance (same as correlation since they are variance 1).
Then $E((z^r_1 - z^r_2)^2) = E(\mu^r_1-\mu^r_2)^2 + 2(1-cov(e_1,e_2)) > 2(1-cov(e_1,e_2))$.

Thus $$cov(e_1,e_2) > 1- 0.5E((z^r_1 - z^r_2)^2)$$ gives a lower bound on the covariance.
The following therefore suggests that the correlation between the two tissues is at least `r 1-0.5*mean((dat$z[,bc]-dat$z[,bch])^2)`.

```{r}
  hist(dat$z[,bc]-dat$z[,bch])
  1-0.5*mean((dat$z[,bc]-dat$z[,bch])^2)
```

But for wb we have a lower bound of `r   1-0.5*mean((dat$z[,bc]-dat$z[,wb])^2)` so we cannot
be sure of any correlation in this case.


## Computations for all tissues

```{r}
  max_absz = apply(abs(dat$z),1, max)
  nullish = which(max_absz < 2)
  nz = dat$z[nullish,]
  vhat = cor(nz)
  saveRDS(vhat,file = "../data/vhat.RDS")
  vhat[bc,bch]
  vhat[bc,wb]
  eigen(vhat)
  dimnames(vhat)=NULL #had to do this to make the figure nice
  corrplot::corrplot(as.matrix(vhat),xlab="")
```



## Session Information

```{r session-info}
```
